
\section{Discussion}

\subsection{Alternative Views}
We address three substantive objections to Agentic Consensus.

\paragraph{``Better models will solve this without a consensus layer.''}
One might argue that sufficiently capable agents will internalize system structure implicitly, making explicit representation unnecessary.
Structurally, the problem is not model capability but human legibility. Even a perfect model that silently maintains a correct internal world model cannot provide the transparency, auditability, and governance that human-AI collaboration requires.
The consensus layer is not a crutch for weak models. It is the interface through which humans exercise meaningful control regardless of model quality.
As models improve, the consensus layer becomes easier to maintain, not unnecessary.
More capable models can produce statistically valid outputs on structurally invalid inputs, and the failure mode becomes harder to detect as scale grows.

\paragraph{``$\Psi$ will hallucinate, making $C$ unreliable.''}
This is a real risk, and we address it by design.
$\Psi$ does not produce a single deterministic rehydrated graph. It produces candidate deltas $\{\Delta C_1, \Delta C_2, \ldots\}$ with uncertainty scores.
Ambiguous rehydration increases $\mathcal{H}(C \mid I)$ and triggers clarification rather than silent reinterpretation.
Round-trip consistency is treated as a soft invariant. The system monitors structural drift $d(C, \Psi(\Phi(C, A)))$ and flags divergence above a threshold rather than demanding exact isomorphism.
Hallucinated structure becomes visible as unsupported claims in $C$ that evidence links fail to corroborate.
The historical failure of UML synchronization tools is instructive but not analogous. Those tools required human maintenance of a secondary artifact. $\Phi$ and $\Psi$ are agent-executed operators running continuously, removing the cost that caused UML to decay.
An escape hatch for low-overhead changes remains a necessary design requirement.

\paragraph{``Graph-based representations are brittle for dynamic, exploratory work.''}
Early-stage exploratory coding is inherently unstructured, and imposing graph schemas too early is a real constraint on creativity. We do not propose that Agentic Consensus applies to all workflows uniformly.
The consensus layer should be incrementally adoptable. Systems begin with partial, low-confidence structure and tighten it as projects stabilize.
High consensus entropy is a signal to not impose structure yet, not a failure.
On cognitive load: $C$ is never exposed as a raw graph. The navigator projects task-relevant subgraphs on demand, so what humans review is a structured summary of the relevant slice, not the full topology.

\subsection{Limitations}
Agentic Consensus raises practical concerns in two areas: system reliability and real-world deployment.

\paragraph{Runtime Risks.}
Agentic Consensus requires careful governance: access control over $C$, redaction for sharing, and escape hatches for low-friction work when structure is unnecessary.
Agents may hallucinate structure; these failures should be visible as high-entropy regions and blocked by required evidence. Evidence itself is not infallible (tests can be flaky, traces noisy, static analyses prone to false positives), so each evidence link in $E$ should carry a confidence weight and contradictory evidence should surface as an explicit inconsistency in $C$, not be silently resolved.

\paragraph{Progressive Adoption.}
For large legacy systems, full rehydration is intractable; $\Psi$ must instead bootstrap incrementally, covering critical modules first and expanding as engineers interact with the system, with uncovered regions marked as high-entropy zones.
For large codebases, $C$ need not be monolithic: sub-graph partitioning along package or service boundaries lets each agent operate on a local view, with the same ownership boundaries that govern human collaboration governing agent coordination.

\subsection{Broader Implications}
Although motivated by coding, Agentic Consensus generalizes to any knowledge discovery system where agents autonomously discover, visualize, and synchronize latent structure in complex datasets or socio-technical systems for human oversight and~intervention.
